---
title: "Nimue Validation Run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nimue Validation Run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 9, 
  fig.height = 9
)
```

Let's compare **safir** to **nimue**. Set up our parameters for running. We use a time step of 0.1 days.

Do a run of **nimue** with increasing vaccination.
```{r}
library(safir)
library(nimue)
library(data.table)
library(individual)
library(ggplot2)
library(dplyr)
library(parallel)

iso3c <- "GBR"
pop <- safir:::get_population(iso3c)
pop$n <- as.integer(pop$n / 1e2)
contact_mat <- squire::get_mixing_matrix(iso3c = iso3c)

tmax <- 365
R0 <- 4

# nimue run
increasing <- run(
  time_period = tmax,
  population = pop$n,
  R0 = R0,
  contact_matrix_set = contact_mat,
  max_vaccine = c(0, seq(10, 1e4, length.out = 40)),
  tt_vaccine = c(0, seq(10, 50, length.out = 40)),
  vaccine_efficacy_disease = rep(0, 17),
  vaccine_efficacy_infection = rep(0.9, 17)
)

```

Do a **safir** run. This takes around 3-4 minutes. That's too long and I'll figure out how to speed it up after it's verified against **nimue**.
```{r}
dt <- 0.1

parameters <- safir::get_parameters(
  population = pop$n,
  contact_matrix_set = contact_mat,
  iso3c = iso3c,
  R0 = R0,
  time_period = tmax
)

parameters <- append_vaccine_nimue(
  parameters = parameters,
  max_vaccine = c(0, seq(10, 1e4, length.out = 40)),
  tt_vaccine = c(0, seq(10, 50, length.out = 40)),
  vaccine_efficacy_disease = rep(0, 17),
  vaccine_efficacy_infection = rep(0.9, 17)
)

timesteps <- parameters$time_period/dt

variables <- create_variables(pop = pop, parameters = parameters)
variables <- create_vaccine_variables_nimue(variables = variables,pop = pop)

events <- create_events(parameters = parameters)
events <- create_events_nimue(events = events,parameters = parameters)

attach_event_listeners(variables = variables,events = events,parameters = parameters, dt = dt)
attach_event_listeners_nimue(variables = variables,events = events,parameters = parameters,dt = dt)

# this is bad and i should feel bad
events$exposure$.listeners[[2]] <- NULL
events$exposure$add_listener(
  safir:::create_exposure_scheduler_listener_nimue(
    events = events,
    variables = variables,
    parameters = parameters,
    dt = dt
  )
)

renderer <- Render$new(timesteps)
vaxx_renderer <- Render$new(timesteps)
processes <- list(
  vaccination_process_nimue(parameters = parameters,variables = variables,events = events,dt = dt),
  infection_process(parameters = parameters,variables = variables,events = events,dt = dt),
  individual::categorical_count_renderer_process(renderer, variables$state, categories = variables$states$get_categories()),
  integer_count_render_process(renderer = vaxx_renderer,variable = variables$vaccine_states,margin = 1:4)
)

setup_events_nimue(parameters = parameters,events = events,variables = variables,dt = dt)

system.time(
  simulation_loop_nimue(
    variables = variables,
    events = events,
    processes = processes,
    timesteps = timesteps,
    progress = FALSE
  )
)
```

Compare vaccinations. These are spot-on.
```{r}
# extract data - vaxx
safir_vax_dt <- as.data.table(vaxx_renderer$to_dataframe())
safir_vax_dt[ , "vaccinated" := X2_count + X3_count ]
safir_vax_dt[ , c("X2_count","X3_count") := NULL]
setnames(x = safir_vax_dt,old = c("X1_count","X4_count"),new = c("unvaccinated","priorvaccinated"))
safir_vax_dt <- melt(safir_vax_dt,id.vars = "timestep",variable.name = "compartment")
safir_vax_dt[, t := timestep * dt]
safir_vax_dt[, timestep := NULL]
safir_vax_dt[, "model" := "safir"]

nimue_vax_dt <- as.data.table(format(increasing, compartments = NULL,summaries = c("unvaccinated","vaccinated","priorvaccinated")))
nimue_vax_dt[, replicate := NULL]
nimue_vax_dt[, "model" := "nimue"]

combined_vax_dt <- rbind(safir_vax_dt,nimue_vax_dt)
```

plot it.
```{r}
ggplot(data = combined_vax_dt) +
  geom_line(aes(x=t,y=value,color=compartment,linetype=model,group=model)) +
  facet_wrap(.~compartment,scales="free_y") +
  theme_bw()
```

Compare state (compartments) output. It's slightly off in deaths, for some reason.
```{r}
# extract data - state
nimue_compartments<- c("S","E","D","R","IMild","ICase","IRec")

df <- renderer$to_dataframe()
safir_dt <- as.data.table(df)
safir_dt[, IMild_count := IMild_count + IAsymp_count]
safir_dt[, IAsymp_count := NULL]
safir_dt <- melt(safir_dt,id.vars = "timestep",variable.name = "compartment",value.name = "value")
safir_dt[, compartment := gsub("(^)(\\w*)(_count)", "\\2", compartment)]
safir_dt[, "t" := timestep * dt]
safir_dt[, timestep := NULL]
safir_dt <- safir_dt[compartment %in% nimue_compartments, ]
safir_dt[ ,"model" := "safir"]

nimue_dt <- as.data.table(format(increasing, compartments = nimue_compartments))
nimue_dt[ , replicate := NULL]
nimue_dt <- nimue_dt[compartment %in% nimue_compartments, ]
nimue_dt[ ,"model" := "nimue"]

combined_dt <- rbind(nimue_dt,safir_dt)
```

plot it.
```{r}
ggplot(data = combined_dt) +
  geom_line(aes(x=t,y=value,color=compartment,group=model,linetype=model)) +
  facet_wrap(.~compartment,scales="free_y")+
  theme_bw()
```
