---
title: "Advanced options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(safir)
```

## Differential modeling of infection and vaccine derived NATs

**safir** allows differential modeling of decay from vaccine and infection derived
neutralizing antibody titre (NAT). To use this feature, set up the model in the same
way as outline in the "Natural Immunity" vignette with the following changes:
  
  1. After all parameters have been made, use `make_independent_vaccine_infection_nat_parameters` and assign the output
  to the same `parameters` object, e.g. `parameters <- make_independent_vaccine_infection_nat_parameters(parameters, ...)`.
  It takes several arguments, `dr_vec_doses` will be a matrix where each column corresponds to a dose, and rows are days.
  The `[i,j]`-th element then corresponds to the decay rate on the `i`-th day since receiving dose `j`. `dr_vec_inf` is a 
  vector giving daily decay for infection derived NAT, and `max_ab_inf` is the maximum value allowed for infection-derived NAT (on natural log scale).
  2. After all variables have been made, call `create_independent_nat_variables` and
  assign the output to the same `variables` object, e.g. `variables <- create_independent_nat_variables(variables = variables, parameters = parameters)`.
  This adds a new `DoubleVariable` object that stores the NAT derived from infection separately.
  3. Instead of using `attach_event_listeners_natural_immunity`, please use `attach_event_listeners_independent_nat`. This makes sure that
  upon recovery from an infection bout, a person's infection-derived NAT boost is added to the variable specifically for infection-derived
  NAT values.
  4. Instead of using `natural_immunity_ab_titre_process`, please use `independent_ab_titre_process`. This calculates decay
  for the two NAT types properly using their specific parameters.

I have listed the specific functions, variables, and parameters that were modified to incorporate
this feature in this [GitHub issue](https://github.com/mrc-ide/safir/issues/64).

This feature will not conflict with the other versions of the model (i.e. it is backwards compatible).
All previous vignettes and examples should still run the way one would expect. If you find a bug
or strange behavior, please make a [GitHub issue](https://github.com/mrc-ide/safir/issues),
and also be sure to check that the two decay parameters and NAT boosts (`mu_ab_infection` in function `make_immune_parameters` and `mu_ab_list` in `get_vaccine_ab_titre_parameters`) are designed such that adding together
the two NAT values prior to calculating efficacy makes sense.

## Time varying parameters

In addition to allowing time varying reproductive numbers via average contact rates, **safir** also supports time varying probabilities of hospitalization, severe disease, and disease outcomes, to support modeling of healthcare changes and variants. 

By default **safir** uses the following parameters from **squire**, and all of them
are vectors where each element gives the relevant probability for each age group.
By default **safir** also uses these vectors, but it also accepts each parameter
as a matrix as well. In this case, the rows correspond to age groups and the columns
correspond to days (not time steps).

1. `parameters$prob_hosp`: P(severe disease | exposure); this probability governs the chance
that the exposed person will require hospitalization.
2. `parameters$prob_asymp`: P(asymptomatic | mild disease); if the exposed person does not need
hospitalization, this probability gives the chance they will be asymptomatic versus mildly symptomatic.
3. `parameters$prob_severe` P(require ICU | hospitalized); if in hospital, does this person require an ICU
bed (i.e. mechanical ventilation)?
4. `parameters$prob_severe_death_treatment` P(death | require ICU, gets bed); if this person
gets an ICU bed, will they die?
5. `parameters$prob_severe_death_no_treatment` P(death | require ICU, don't get bed); if this
person doesn't get an ICU bed, will they die?
6. `parameters$prob_non_severe_death_treatment` P(death | require hosp, get bed); if in hospital
but does not require ICU bed, will this person die?
7. `parameters$prob_non_severe_death_no_treatment` P(death | require hosp, don't get bed); if needs a hospital
(but not ICU bed) but does not recieve one, will this person die?
