---
title: "Natural Immunity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{natural_immunity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(safir)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(safir)
library(individual)
library(nimue)
library(data.table)
library(ggplot2)
```

# Setup

This vignette is essentially identical to the one titled "Vaccine model with multiple doses", but includes options for modeling immune response mechanistically (a boost to antibody titre as the result of natural infection) as opposed to with an R compartment within which agents are 100% immune for re-infection. The only thing that needs to be added to the `parameters` list is `parameters$mu_ab_infection`, giving the antibody boost derived from each natural infection. It is a vector where the 1st element is the antibody boost from someone's first infection, 2nd from someone's second subsequent infection, and so on. If an individual is getting reinfected a number of times greater than the length of the vector, the last element of the vector is used as their antibody titre associated with that reinfection event. 

```{r}
iso3c <- "GBR"
pop <- safir::get_population(iso3c)
pop$n <- as.integer(pop$n / 1e3)
contact_mat <- squire::get_mixing_matrix(iso3c = iso3c)

tmax <- 100
dt <- 0.5
R0 <- 4

# vaccine dosing
vaccine_doses <- 2
dose_period <- c(NaN, 28)
vaccine_set <- c(0, seq(from = 1e3, to = 1e4, length.out = tmax-1))
vaccine_set <- floor(vaccine_set)

# vaccine strategy
vaccine_coverage_mat <- nimue::strategy_matrix(strategy = "Elderly",max_coverage = 0.2)
next_dose_priority <- matrix(data = 0, nrow = vaccine_doses - 1,ncol = ncol(vaccine_coverage_mat))
next_dose_priority[1, 15:17] <- 1 # prioritize 3 oldest age groups for next dose

# base parameters
parameters <- safir::get_parameters(
  population = pop$n,
  contact_matrix_set = contact_mat,
  iso3c = iso3c,
  R0 = R0,
  time_period = tmax,
  dt = dt
)

# vaccine parameters
ab_parameters <- get_vaccine_ab_titre_parameters(vaccine = "Pfizer", max_dose = vaccine_doses,correlated = FALSE)

# combine parameters and verify
parameters <- make_vaccine_parameters(
  safir_parameters = parameters,
  vaccine_ab_parameters = ab_parameters,
  vaccine_set = vaccine_set,
  dose_period = dose_period,
  strategy_matrix = vaccine_coverage_mat,
  next_dose_priority_matrix = next_dose_priority
)

# ab boost for each infection
parameters$mu_ab_infection <- ab_parameters$mu_ab
```

# Run **safir**

This part is also exactly the same as the "Vaccine model with multiple doses" except the following places:

  1. `create_natural_immunity_variables` is called to attach additional variables to the `variables` list needed to model multiple infections
  2. `attach_event_listeners_natural_immunity` is called to add additional event listeners to `events` list needd to antibody boosting. Please note there is an argument `additive`
  which can be used to change how antibody boosting from infection interacts with existing antibody titre levels, so please carefully read the documentation for `attach_event_listeners_natural_immunity()`
  3. We use another render object, `inf_renderer` which will plot the number of cumulative infections the population has. Rendering adds more computation,
  increasing model run time, so for actual use, please don't add this and its associated process function `integer_count_render_process_daily` unless you are interested
  in tracking this information.
  4. `natural_immunity_ab_titre_process` is added to the `processes` list to update antibody titre from vaccines or natural immunity (replaces `vaccine_ab_titre_process`).

As usual, for actual use, please use only the minimum number of output/rendering needed for your analysis, as it increases run time. In
particular, `double_count_render_process_daily` to output the antibody titre for each person each day is time consuming.

```{r}
# create variables
timesteps <- parameters$time_period/dt
variables <- create_variables(pop = pop, parameters = parameters)
variables <- create_vaccine_variables(variables = variables,parameters = parameters)
variables <- create_natural_immunity_variables(variables = variables, parameters = parameters)

# create events
events <- create_events(parameters = parameters)
events <- create_events_vaccination(events = events,parameters = parameters)
attach_event_listeners(variables = variables,events = events,parameters = parameters, dt = dt)
attach_event_listeners_vaccination(variables = variables,events = events,parameters = parameters,dt = dt)
attach_event_listeners_natural_immunity(variables = variables, events = events, parameters = parameters, dt = dt)

# make renderers
renderer <- Render$new(parameters$time_period)
ab_renderer <- matrix(data = NaN,nrow = parameters$time_period,ncol = sum(parameters$population))
dose_renderer <- Render$new(parameters$time_period)

double_count_render_process_daily <- function(variable, dt) {
  stopifnot(inherits(variable, "DoubleVariable"))
  function(t) {
    if ((t * dt) %% 1 == 0) {
      ab_renderer[as.integer(t * dt), ] <<- variable$get_values()
    }
  }
}

inf_renderer <- Render$new(parameters$time_period)

# processes
processes <- list(
  natural_immunity_ab_titre_process(parameters = parameters,variables = variables, dt = dt),
  vaccination_process(parameters = parameters,variables = variables,events = events,dt = dt),
  infection_process_vaccine_cpp(parameters = parameters,variables = variables,events = events,dt = dt),
  categorical_count_renderer_process_daily(renderer = renderer,variable = variables$states,categories = variables$states$get_categories(),dt = dt),
  double_count_render_process_daily(variable = variables$ab_titre,dt = dt),
  integer_count_render_process_daily(renderer = dose_renderer,variable = variables$dose_num,margin = 0:vaccine_doses,dt = dt),
  integer_count_render_process_daily(renderer = inf_renderer,variable = variables$inf_num,margin = 0:51,dt = dt)
)

setup_events(parameters = parameters,events = events,variables = variables,dt = dt)
```

Run the simulation.

```{r}
system.time(simulation_loop_safir(
    variables = variables,
    events = events,
    processes = processes,
    timesteps = timesteps,
    variables_dont_update = c("discrete_age", "phase"),
    progress = FALSE
))
```

## Plot Results

### Antibody titre

This lets us check if people are getting vaccinated; Ab titre, averaged over the population, relative to the time of the first dose. It won't look like the cohort plots because everybody gets their second dose at slightly different times, depending on availability.

```{r}
# plot: ab titre
vaccinated_or_infected <- variables$dose_num$get_index_of(set = 0)
vaccinated_or_infected$not(inplace = TRUE)
vaccinated_or_infected$or(variables$inf_num$get_index_of(set = 0)$not(inplace = TRUE))

ab_titre <- ab_renderer[, vaccinated_or_infected$to_vector()]
ab_titre[which(!is.finite(ab_titre))] <- NaN
cols_2_drop <- which(is.nan(ab_titre[nrow(ab_titre), ]))
if (length(cols_2_drop) > 0) {
  ab_titre <- ab_titre[, -cols_2_drop]
}
start <- apply(ab_titre, 2, function(x){ which(abs(x - 0) > 2e-7)[1] })

ab_titre <- lapply(X = 1:ncol(ab_titre),FUN = function(x){
  ab_titre[start[x]:nrow(ab_titre) , x]
})

ab_titre_dt <- data.table(
  ab = unlist(ab_titre),
  t = unlist(lapply(ab_titre, function(ab_titre){1:length(ab_titre)})),
  id = rep(1:length(ab_titre), times = vapply(ab_titre,length,integer(1)))
)

ab_titre_quant_dt <- ab_titre_dt[, .(mean = mean(ab), lo = quantile(ab,probs =0.025) , hi = quantile(ab,probs = 0.975)) , by = .(t)]

ggplot(data = ab_titre_quant_dt) +
  geom_line(aes(x=t,y=mean)) +
  geom_ribbon(aes(x=t,ymin=lo,ymax=hi),alpha=0.5) +
  theme_bw()

```

### Proportion of population with each dose

We also want to plot the number of people with each dose over time to check it's working.

```{r}
dose_out <- dose_renderer$to_dataframe()
colnames(dose_out)[2:(vaccine_doses+2)] <- as.character(0:vaccine_doses)
dose_out <- melt(as.data.table(dose_out),id.vars="timestep")
setnames(dose_out, "variable", "dose")

ggplot(data = dose_out) +
  geom_line(aes(x=timestep,y=value,color=dose)) +
  theme_bw()
```

### Infection states

The compartmental states of the model.

```{r}
saf_dt <- as.data.table(renderer$to_dataframe())
saf_dt[, IMild_count := IMild_count + IAsymp_count]
saf_dt[, IAsymp_count := NULL]
saf_dt <- melt(saf_dt,id.vars = c("timestep"),variable.name = "name")
saf_dt[, name := gsub("(^)(\\w*)(_count)", "\\2", name)]
setnames(x = saf_dt,old = c("timestep","name","value"),new = c("t","compartment","y"))

ggplot(data = saf_dt, aes(t,y,color = compartment)) +
  geom_line() +
  geom_line() +
  facet_wrap(~compartment, scales = "free")
```

### Cumulative infections

Here we plot how many people in the population have had 0,1,2,... infections.

```{R}
# infections
inf_out <- inf_renderer$to_dataframe()
inf_out <- inf_out[, -which(sapply(inf_out, function(x){all(x == 0)}, USE.NAMES = FALSE))]
inf_out <- as.data.table(inf_out)
inf_out <- melt(inf_out, id.vars = c("timestep"),variable.name = "infections")
inf_out[, infections := gsub("(^X)(\\w*)(_count)", "\\2", infections)]
inf_out[, t := timestep * dt]
inf_out[, timestep := NULL]

ggplot(data = inf_out, aes(t,value,color = infections)) +
  geom_line(size = 1.25, alpha = 0.75)
```

# Code design

The new code is in:

  1. R/efficacy_naturalimmunity.R: contains `natural_immunity_ab_titre_process`, `get_time_since_last_dose_or_infection` (helper function)
  2. R/events_naturalimmunity.R: contains `attach_event_listeners_natural_immunity`
  3. R/variables_naturalimmunity.R:  contains `create_natural_immunity_variables`
